---
editor_options: 
  markdown: 
    wrap: 72
---

# Exercises

## Exercise 1 & 2 {.unnumbered}

```{=html}
<script type="text/x-mathjax-config">
MathJax.Hub.Register.StartupHook("TeX Jax Ready",function () {
  MathJax.Hub.Insert(MathJax.InputJax.TeX.Definitions.macros,{
    cancel: ["Extension","cancel"],
    bcancel: ["Extension","cancel"],
    xcancel: ["Extension","cancel"],
    cancelto: ["Extension","cancel"]
  });
});
</script>
```
> ***Math Prerequisite:*** Integration

### The biology

Here we work with a very important concept, individual growth. If one
understands how an organism grows, one can begin to see how growth
informs everything from population biomass to maturity. The equation we
are working with is a famous one, the one presented by Ludwig von
Bertalanffy. While the growth equation has stood the test of time and
bears his name, it's important to point out that Ludwig was a literal
nazi-party member.

That being said, we start off by simply expressing that the weight of an
individual at a given age $W(t)$ is related to its length at the same
age $L(t)$ to the third power multiplied by the density $\rho$ of the
organism resulting in $W(t) = \rho L(t)^3$. Most organisms but not all
follow this 3-dimensional simplified view so the third power makes
intuitive sense. We now step into the logic of the growth equation. If
we assume that an organism's change in weight over time is dependent on
anabolic (using energy and mass to build, muscle building) and catabolic
(breaking down mass to produce energy, digestion) processes we show that
simply as the following.

$$
\\[10pt]
\frac{dW}{dt} = \text{anabolic - catabolic}
\\[10pt]
$$ We take things a step further and translate the anabolic and
catabolic to quantifiable terms that link to an organism's metabolism.
Here Marc explains the justification for surface area for the anabolic
term and volume for the catabolic term. It's important to realize that
these are assumptions and can be changed accordingly.

$$
\\[10pt]
\frac{dW}{dt} = \sigma L^2 -cL^3
\\[10pt]
$$ The next step is to incorporate our previous assumption of the
3-dimensional organism. Since we know that change in weight over time is
related to the volume of the organism, one of the "L"'s has to be
dependent on time also, we get
$\frac{dW}{dt} = 3\rho L(t)^2 \frac{dL}{dt}$ by taking the derivative.
Now its just substitution to include this new length view of growth into
our metabolic view of growth. And lastly, Marc then rearranges the
equation to show the second form of the von Bertalanffy growth equation
which for all practical matters is much easier to interpret.

$$
\\[10pt]
\frac{dW}{dt} = 3\rho L^2 \frac{dL}{dt} = \sigma L^2 -cL^3
\\[10pt]
\text{after simplifying we get}
\\[10pt]
\frac{dL}{dt} = \frac{\sigma}{3\rho}-\frac{c}{3\rho}L
\\[10pt]
\text{after rearranging we get}
\\[10pt]
\frac{dL}{dt} = \frac{c}{3\rho} \left(\frac{\sigma}{c}-L\right) = k(L_{\infty}-L)
$$

The rearranged equation tells us that change in length given an age
corresponds to some $k$ growth component multiplied by the difference
between the largest that organism can be with its actual length!

### The challenge

<center>

**Check that the units of q, k, and asymptotic size are correct and find
a unique solution for the ordinary differential equation given**
$L(0)=L_0$

</center>

### The solution

Here Marc suggested an alternative way to group the parameters and we
need to check the units so they make sense. $$
\\[10pt]
q = \frac{\sigma}{3\rho} 
\\[10pt]
k = \frac{c}{3\rho}
\\[10pt] 
\frac{dL}{dt} = q-kL
$$ If the derivative is some unit length over time then we can assume
something like centimeters per day. That would require the other two
terms to have the same units. $q$ is by itself so centimeters per day is
fine but $k$ is a little trickier. We know $L$ is centimeters but we
can't have two lengths multiplied together otherwise we would get
centimeters squared. That would suggest that $k$ is just some unitless
number per day.

*checking our work* : Recall that $\rho$ was a density so that means it
has to have units of weight divided by volume such as $\frac{kg}{cm^3}$.
With simple substitution cancelling we can see that the units for
$q = \frac{\sigma}{\rho} = \frac{cm}{day}$. And similarly, if we use the
units from $c$ which is just the same units as sigma but one power
higher, then our units for $k$ are also correct when we substitute and
cancel.

Lastly, referring to the other form of the equation with $L_\inf$ we
need to figure out those units. The units for $\sigma$ come directly
from the book and we just figured out what $c$'s units. So when we
multiply the two to get asymptotic length ($L_inf$), most of the units
cancel only leaving cm.

$$
\\[10pt]
L_\inf = \frac{\sigma}{c} 
\\[10pt]
\sigma = \frac{kg}{day*cm^2}
\\[10pt] 
c = \frac{kg}{day*cm^3}
\\[10pt]
\frac{\sigma}{c} = cm
$$ Now to find the unique solution for the equation that will require
some integration.

$$
\text{we start with our equation}
\\[10pt]
\frac{dL}{dt} = k(L_{\infty}-L)\\[10pt]
\\[10pt] 
\text{we then try to group all the L's together}
\\[10pt]
\frac{dL}{(L_{\infty}-L)} = k*dt
\\[10pt]
\text{now this is simpler to integrate}
\\[10pt]
\int{\frac{dL}{(L_{\infty}-L)}}=k\int{dt}
\\[10pt]
\text{notice we moved k out front and now we integrate}
\\[10pt]
$$ Now remember that the derivative of the natural log of a number is
one over that number. Since we want the integral we just need to go in
reverse. If you notice the left side of the equation is one over a
number. Our integral will be a natural log! And for the other side it's
just $t$ plus some constant.

$$
\text{integrate using natural log rule, notice the negative out front}
\\[10pt]
-ln(L_\infty-L(t))=kt+C
\\[10pt] 
\text{for simplicity let's move the negative over}
\\[10pt]
ln(L_\infty-L(t))=-kt+C
\\[10pt]
\text{now we want to remove the ln by using an exponent which applies to both sides but not C}
\\[10pt]
exp(ln(L_\infty-L(t)))=C*exp(-kt)
\\[10pt]
\text{which after simplifying and isolating L(t), looks like this}
\\[10pt]
L(t)=L_\infty-C*exp(-kt)
\\[10pt]
\text{now that we have L(t) we can use the initial condition Marc provided, where L(0) = L0, t = 0, to figure out what C is}
\\[10pt]
L(t)=L_\infty-C*exp(-k*0) = L_0
\\[10pt]
C = L_\infty-L_0
\text{Now we just plug C back into the equation to figure out L(t) with a real constant!}
\\[10pt]
L(t)=L_\infty-(L_\infty-L_0)*exp(-kt)
\\[10pt]
\text{distribute the new C value}
\\[10pt]
L(t)=L_\infty-L_\infty*exp(-kt)+L_0*exp(-kt)
\\[10pt]
\text{factor out Linf}
\\[10pt]
L(t)=L_\infty(1-exp(-kt)+L_0*exp(-kt)
$$

### Checking our work

Since this is the equation told us to work to, no need to check work
here.

### Reflecting on the biology

This growth equation is used extensively in fisheries and many other
applications. If you were to obtain a reliable estimate of growth for an
organism you can learn a lot about its life such as when does maturity
likely occur or the maximum size you would expect to see in the wild.

Here is what the growth curve looks plotted out if we assume $L_0$ is 0.
The result shows a variety of fish that live up to 25 years old
(assuming they make it that far before being eaten), with a variety of
$L_\infty$ values. Now the tricky part after doing the math is catching
the critters and asking how old they are when you measure them!

```{r}
vbg <- function(k,Linf,x) Linf*(1-exp(-k*x))
Linf <- seq(50,110,10)
col<-rainbow(length(Linf))

for(i in seq_along(Linf)) {
    curve(vbg(.15, Linf[i], x), from=0, to=25, add=i!=1, col=col[i], ylim=c(0,130), ylab = "Length (cm)", xlab = "Age (y)", main="Multiple values of L_inf with a k=0.15")
}

```

## Exercise 3 {.unnumbered}

> ***Math Prerequisite:*** derivatives

### The biology

If we recall that $b$ is our allometric parameter and it is often around
3 for fish we can plug this into the maturity equation provided.

$$
t_m = \frac{1}{k}*log\left(\frac{m + bk}{m}\right)
$$

Similarly, we can use the three different mortality values that Marc
provided (0.1, 0.2, 0.5) in combination with a sequence of $k$, growth
rate values, to visualize optimal maturity ages.

```{r}
tm <- function(b,m,k) 1/k * log((m + b*k)/m)
agem <- c(0.1, 0.2, 0.5)
col<-rainbow(length(agem))

for(i in seq_along(agem)) {
    curve(tm(3, agem[i], x), from=0.1, to=2, add=i!=1, col=col[i], ylim=c(0,15), ylab = "Optimal age at maturity", xlab = "Growth rate, k", main="Multiple values of M with b=3")
}
```

The main point here is to realize that in systems where natural
mortality is high, whether this is disease, predation, starvation or any
other source, they incentive an earlier maturity age. This makes
intuitive sense, if you are going to die younger, you should have kids
younger. But there's also another important point, the strength of this
relationship decreases as growth rate tends to increase and also reduces
age at maturity.

But how did we come to this conclusion? This is what Marc is asking us
to do now. Show that this equation is possible from first principles by
using our understanding of growth and an exponential distribution for
survival. And specifically find the length at maturity.

### The challenge

<center>

**Find length at maturity assuming that fitness could be represented
by** $F(t) = e^{-mt}*f*L(t)^b$

</center>

### The solution

In the previous exercise, Marc assumed that fitness could be represented
by $F(t) = e^{-mt}\cdot f \cdot L(t)^b$ where the first half of the term
corresponds to survival given an age, and the second half of the term
corresponds to reproductive output given an age and scaling component.
The first step is to replace $L(t)$ with our previously derived growth
equation which would look like something like this. But we can also do
one extra step to make things easier for ourselves, we cancel out the
"birth" term, where we assume size is equal to 0. Realize this
assumption may not be appropriate if size at birth is relatively large
compared to $L_\infty$. Also, fecundity, f is a constant so we can set
that to 1.

$$
F(t) = e^{-mt} \cdot f (L\_\infty(1-e^{(-kt)}+L_0*exp(-kt))^b

\\[10pt] 
\text{for simplicity we can drop the last term and set f = 1}
\\[10pt]

F(t) = e^{-mt} \cdot 1 (L\_\infty(1-e^{(-kt)}))^b
\cancel{+L_0*exp(-kt))} 
$$

Since we want to figure out the change in
$F$ given this function we take the derivative of this with respect to
$L$.

$$

F(t) = e^{-mt} \cdot 1 (L_\infty(1-e^{(-kt)}))^b

\\[10pt] 
\text{distribute L infinity and take derivative}
\\[10pt]


\frac{d}{dt}\left(e^{-m t}\left(l_\infty-l_\infty e^{-k t}\right)^{b}\right)


\\[10pt] 
\text{Use the product rule for the survival and reproductive output}
\\[10pt]


=\left(l_\infty-e^{-k t} l_\infty\right)^{b}\left(\frac{d}{d t}\left(e^{-m t}\right)\right)+e^{-m t}\left(\frac{d}{d t}\left(\left(l_\infty-e^{-k t} l_\infty\right)^{b}\right)\right)


\\[10pt] 
\text{Using the chain rule for the first part of our product rule}
\\[10pt]


=\left(l_\infty-e^{-k t} l_\infty\right)^{b} e^{-m t}\left(\frac{d}{d t}(-m t)\right) + e^{-m t}\left(\frac{d}{d t}\left(\left(l_\infty-e^{-k t} l_\infty\right)^{b}\right)\right)

\\[10pt] 
\text{Factor out constant, m, and derivative of t = 1}
\\[10pt] 


=\left(l_\infty-e^{-k t} l_\infty\right)^{b} e^{-m t}\cdot-m + e^{-m t}\left(\frac{d}{d t}\left(\left(l_\infty-e^{-k t} l_\infty\right)^{b}\right)\right)

\\[10pt] 
\text{Now let's finish the other half of our original product rule by using the chain rule again}
\\[10pt]

=-e^{-m t} m\left(l_\infty-e^{-k t} l_\infty\right)^{b}+b \cdot  e^{-m t}\left(l_\infty-e^{-k t} l_\infty\right)^{b-1}\left(\frac{d}{d t}\left(l_\infty-e^{-k t} l_\infty\right)\right)

\\[10pt]
\text{take the derivative of each inside term}
\\[10pt]

\begin{aligned}
&=-e^{-m t} m\left(l_\infty-e^{-k t} l_\infty\right)^{b}+ -l_\infty\left(\frac{d}{d t}\left(e^{-k t}\right)\right)+\frac{d}{d t}\left(l_\infty\right) \cdot b \cdot e^{-m t}\left(l_\infty-e^{-k t} l_\infty\right)^{b-1}
\end{aligned}


\\[10pt]
\text{ use the chain rule for the first derivative, factor out constant, k, and derivative of t = 1}
\\[10pt]


\begin{aligned}
=&-e^{-m t} m\left(l_\infty-e^{-k t} l_\infty\right)^{b}+ \left(  l_\infty \cdot k \cdot e^{-k t}  + \frac{d}{d t}(l_\infty)\right) \cdot b\cdot e^{-m t}\left(l_\infty-e^{-k t} l_\infty\right)^{b-1}
\end{aligned}

\\[10pt]
\text{ The derivative of L infinity is zero }
\\[10pt] 


=-e^{-m t} m\left(l_\infty-e^{-k t} l_\infty\right)^{b} + \left(l_\infty \cdot k \cdot e^{-k t} +0\right) \cdot b\cdot e^{-m t}\left(l_\infty-e^{-k t} l_\infty\right)^{b-1}

\\[10pt]
\text{ So the final equation is the following }
\\[10pt] 

F'(t) = -e^{-m t} m\left(l_\infty-e^{-k t} l_\infty\right)^{b} + \left(l_\infty \cdot k \cdot e^{-k t}\right) \cdot b\cdot e^{-m t}\left(l_\infty-e^{-k t} l_\infty\right)^{b-1}

\\[10pt]
\text{ which we can flip around and simplify to look nice}
\\[10pt] 

F'(t) = 
b\cdot e^{-kt-mt} \cdot l_\infty \cdot k\left(l_\infty-e^{-k t} l_\infty\right)^{b-1}
-e^{-m t} m\left(l_\infty-e^{-k t} l_\infty\right)^{b}

$$
That was a beast of a derivative. Let's take a moment to recap what we did. We started with a function that describes fitness over time now we are looking at how fitness changes over time. To find the optimum length of maturity we still need to solve for $t$ after setting the equation equal to 0. 




$$

0 = b\cdot e^{-kt-mt} \cdot l_\infty \cdot k\left(l_\infty-e^{-k t} l_\infty\right)^{b-1}
-e^{-m t} m\left(l_\infty-e^{-k t} l_\infty\right)^{b}

\\[10pt]
\text{this is a little complex so lets rename all coefficeints}
\\
\text{G in the first term and S in the second term and seperate the exponent}
\\[10pt] 

0 = \frac{G \left(l_\infty-e^{-k t} l_\infty\right)^{b}}{\left(l_\infty-e^{-k t} l_\infty\right)}-S\left(l_\infty-e^{-k t} l_\infty\right)^{b}

\\
\text{now we can divide out the parantheses with the b to get}
\\[10pt] 

0 = \frac{G}{\left(l_\infty-e^{-k t} l_\infty\right)} - S


\\[10pt] 
\text{we can also divide by G to get}
\\[10pt] 

0 = \frac{1}{\left(l_\infty-e^{-k t} l_\infty\right)} - \frac{S}{G}

\\[10pt] 
\text{now recall G and S are placeholders but now alot of these coefficients (L infinity) can cancel out and simplify}
\\[10pt] 

0 = \frac{1}{\left(1-e^{-k t}\right)} - \frac{me^{kt}}{bk}

\\[10pt] 
\text{we can now move the 2nd term over}
\\[10pt] 

 \frac{1}{\left(1-e^{-k t}\right)} = \frac{me^{kt}}{bk}

\\[10pt] 
\text{we can multiply out the bottom parantheses and distribute it to the other term}
\\[10pt] 

1 = \frac{me^{kt}}{bk} - \frac{m}{bk}

\\[10pt] 
\text{great! now m, b, and k can move over}
\\[10pt]

\frac{bk}{m} = e^{kt} - 1

\\[10pt] 
\text{add the one and take the log to remove the exponent}
\\[10pt]

log\left(\frac{bk}{m} + 1\right) = kt


\\[10pt] 
\text{now just move the k over}
\\[10pt]

\frac{1}{k}log\left(\frac{bk}{m} + 1\right) = t

\\[10pt] 
\text{and the inside just simplifies to m as denominator}
\\[10pt]

t = \frac{1}{k}*log\left(\frac{m + bk}{m}\right)

$$


### Checking our work

No need to check our work here! Our equation matches Marc's!

### Reflecting on the biology
